---
title: "Circular Dataset Structures"
output: html_document
date: "`r Sys.Date()`"
author: "James Foster"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Details

> **AUTHOR**
>
> James Foster

> **DESCRIPTION**
>
> Examination of different dataset structures that may arise in the study of animal navigation.

> **INPUTS**
>
> `unwrap_functions.R`

> **OUTPUTS**
>
> Plotted results.

> **REFERENCES**
>
> **data types**
>
> Duelli P. & Wehner R.
> (1973) The Spectral Sensitivity of Polarized Light Orientation in Cataglyphis bicolor ( Formicidae , Hymenoptera ) *Journal of Comparative Physiology* 53(3) 37-53
>
> Edrich, W., Neumeyer, C.
> and von Helversen, O.
> (1979).
> “Anti-sun orientation” of bees with regard to a field of ultraviolet light.
> *J. Comp. Physiol.* 134, 151–157.
>
> **modelling methods**
>
> Sayin S, Couzin-Fuchs E, Petelski I, Günzel Y, Salahshour M, Lee CY, Graving JM, Li L, Deussen O, Sword GA, et al. (2025) The behavioral mechanisms governing collective motion in swarming locusts.
> Science.
> 387(6737):995–791
>
> Gabry J, Češnovar R, Johnson A (2022).
> cmdstanr: R Interface to 'CmdStan'.
> <https://mc-stan.org/cmdstanr/>
>
> Bürkner, P.-C.
> (2018).
> Advanced Bayesian Multilevel Modeling with the R Package brms.
> The R Journal 10, 395–411.
>
> Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M., Guo, J., Li, P. and Riddell, A.
> (2017).
> Stan: A Probabilistic Programming Language.
> Journal of Statistical Software 76 doi: 10.18637/jss.v076.i01

> **TODO**
>
> -   move necessary functions to unwrap_functions.R
> -   make functions for result extraction from BRMS

## Set up workspace

```{r Load functions and packages}
source('unwrap_functions.R')
```

Set up some colours for plotting.

```{r Set up plot colours}
col_kappa = '#1E78B5'#colour for kappa parameter
col_rho = '#F08024' #colour for mean vector length
col_sd = '#E74A29' #colour for SD and mean angle
col_sd2 = '#E57461' #colour for other SD heuristics
col_pdf = adjustcolor(col = '#21A885', # colour for probability density
                      alpha.f = 0.7)
col_obs = '#3E1F51' #colour for control observations
col_treat = '#006400' # colour for treatment observations
```

# Divergence from home direction

In navigation experiments return directions are often compared with the expected home direction to determine if homeward navigation is perturbed.
Here, we simulate example data (𝑛 = 20) generated from a distribution with a mean differs that from the expected direction (0°) by only 15°.

```{r Generate diverging data, echo  = FALSE}
ndata = 20 #sample size

# circular zero expected angle
c0 = circular(x = 0,
              units = 'degrees',
              rotation = 'clock',
              zero = pi/2)
#true mean angle
c15 = circular(x = -15,
              units = 'degrees',
              rotation = 'clock',
              zero = pi/2)

par(pty = 's') # square axes
par(mar = c(0,0,0,0)) # no margins

cd_divergence = DescriptCplot(m = -15, # population mean
                              k = 10, # high concentration
                              refline = 0, # expected direction
                              ndata = ndata,
                              mvcol = col_obs, #distr. mean same colour as observations
                              sdcol = NA, # don't plot SD
                              denscol = NA, # don't plot prob. density
                              save_sample = TRUE)
```

A common shortcut to assess orientation in an expected direction is to use the v-test modification of the Rayleigh test for uniformity.
This tests the hypothesis of uniformity against the alternative of clustering in some direction near the expected direction.
The data are very far from uniform ($\kappa = 10$), and the mean angle is very close to the expected angle (cosine similarity of `cos(rad(-1)) = 97%`), so the test is significant.
Even so, the data show a trend away from the expected angle, 75% of points falling to the left of 0°.

```{r v-test for uniformity}
#v-test finds significant orientation in 0°, but we know the true direction is different
rayleigh.test(cd_divergence, mu = c0)

```

## Fit the model

In order to fit a circular model using the shifted modulo and softplus links, we need to define a custom family.
This family

```{r Set up the circular model custom family}
#set up required Stan functions
#shifted modulo in Stan code
mod_circular_fun = stanvar(scode = "
  real mod_circular(real y) {
    return fmod(y + pi(), 2*pi()) - pi();
  }
",
                           block = 'functions')

#custom likelihood function using the shifted modulo link
stan_unwrap_fun = stanvar(scode = "
  real unwrap_von_mises_lpdf(real y, real mu, real kappa) {
    return von_mises_lpdf(y | mod_circular(mu), kappa);
  }
  real unwrap_von_mises_rng(real mu, real kappa) {
    return von_mises_rng( mod_circular(mu) , kappa);
  }
",
                          block = 'functions') 
#define the custom family
unwrap_von_mises = custom_family(
  name = "unwrap_von_mises",
  dpars = c("mu",
            "kappa"),
  links = c('identity',#N.B. the link function is defined via the LPD function
            "softplus"), 
  lb = c(-pi,#lower bound of mu should be -pi
         0),
  ub = c(pi,#upper bound of mu should be pi
         NA),
  type = "real",#takes continuous response data
  # vars = stan_unwrap_fun + mod_circular_fun
)

```

To fit out model, we need to convert our data from degrees to radians.
It is also recommended to remove additional formatting by converting from class "circular" to class "numeric".
Our formula indicates that each of mu and kappa require a single parameter estimate (\~1).
In this data, we expect a mean direction anywhere on the circle (within $\pi$ of the mean) and high concentration ($\kappa>1$), so highest prior density is set in that range.
Since the mean direction is far from the wrap-point ($-\pi$ or $\pi$), the `Rhat` convergence heuristic for the mu `Intercept` can be interpreted on the linear scale.

```{r Fit a model to the divergence data}
#reformat data for BRMS
dt_divergence = data.frame(y = as.numeric(#remove circular formatting
                                rad(cd_divergence)#convert to radians
                                )
                           )#make a data frame

#formula for a circular model with no predictors
form_divergence = bf(y~1,
                     kappa~1,
                     family = unwrap_von_mises)
#unbiased priors
prior_divergence =  prior('normal(0,2*pi())', class = 'Intercept', dpar = 'mu') +
                     prior('normal(3,3)', class = 'Intercept', dpar = 'kappa')

#fit a generic unwrap model
mod_divergence = brm(formula = form_divergence,
             data = dt_divergence,
             family = unwrap_von_mises,
             stanvars = stan_unwrap_fun + mod_circular_fun,
             prior = prior_divergence,
             silent = 2,
             backend = 'cmdstan'#faster and more reliable
  )

summary(mod_divergence)
```

A model fitted to the example data correctly indicates that the mean direction is to the left of the expected direction in \>99% of 𝜇 estimates.

```{r Plot the model fitted for divergence data}
draws_divergence = as_draws_df(mod_divergence)


par(pty = 's')
par(mar = c(0,0,0,0),
    mfrow = c(1,2))
PCfun(cd_divergence,
      col = col_obs,
      sep = 0.05,
      shrink = 1.25,
      plot_rho = FALSE)
Draws2Cont(draws_divergence,
           x_string = )
arrows.circular(x = circular(-15,
                             units = 'degrees',
                             rotation = 'clock',
                             zero = pi/2),
                y = A1(10),
                col = col_obs,
                lwd = 5,
                length = 0.1/1.25
)


with(draws_divergence,
     arrows.circular(x = median.circular(x =
                                           circular(Intercept,
                                  rotation = 'clock',
                                  zero = pi/2)
                                  ),
                     y = A1(softplus(x = median(x = Intercept_kappa))),
                     lwd = 2,
                     length = 0.1/1.25,
                     col = adjustcolor(col_sd, alpha.f = 200/255))
)
with(draws_divergence,
     {
      VertHist(data = deg(Intercept), 
               main = '\nmean angle',
               ylim = c(-30, 15),
               col = adjustcolor(col_sd, alpha.f = 100/255),
               cex.axis = 0.7)
     }
)
abline(h = 0,
       col = 'gray',
       lwd = 7)

with(draws_divergence,
     {
       paste0('mu left of expected angle: ',
              mean(Intercept < 0)*100, '%')
      }
     ) #nearly all estimates suggest a rightwards turn
```
## Test the expected angle hypothesis
To test the hypothesis that the true mean direction falls in the expected direction, we can fit another model where this is explicitly stated. To inform BRMS that we do not want to estimate the mean direction, we use the notation `y~0`, which removes the intercept. We can now estimate $\kappa$, _assuming_ that the mean direction is 0°, using the same prior for kappa as before.
The model for a divergent mean direction has higher expected log predictive density (ELPD) than the model for the expected direction. This ELPD difference is $>2\times$ larger than its standard error, so we can state with some confidence that the hypothesis of a divergence from the expected direction is more likely than the hypothesis of alignment with the expected direction.
```{r Model comparison observed vs expected mean direction}


#fit a model with mean at 0
mod_expect = brm(formula = bf(y~0,
                              kappa~1),
                data = dt_divergence,
                family = unwrap_von_mises,
                prior = prior('normal(3,3)', class = 'Intercept', dpar = 'kappa'),
                stanvars = stan_unwrap_fun + mod_circular_fun,
                silent = 2,
                backend = 'cmdstan'
                )
#add Leave-One-Out Cross-Validation

loo_divergence = loo(mod_divergence)
loo_expected = loo(mod_expect)
loo_compare(loo_divergence, loo_expected)
```

In this case, the expected angle model is straightforward to fit because the expected angle is 0°, which is also the default intercept for a `y~0` (no-intercept) model.
If the expected angle were not 0°, we would need to adjust the data to align the expected angle with the `0` intercept.
Here is an example with the expected angle at -15°.

```{r Model comparison for expected angle not at 0}
#align the data relative to the expected angle
dt_aligned = within(dt_divergence,
                    {
                    y = y - rad(-15)  #subtract the expected angle
                    }
                    )


#we now place our expected angle at -15°, the true mean
mod_true = brm(formula = bf(y ~ 0,
                            kappa~1),
               data = dt_aligned, 
               family = unwrap_von_mises,
               prior = prior_divergence[2,], # just the kappa prior
               stanvars = stan_unwrap_fun + mod_circular_fun,
               silent = 2,
               backend = 'cmdstan'
)

#this model should be compared with another fitted to the
#_same data_
mod_false =  brm(formula = bf(y ~ 1,
                            kappa~1),
               data = dt_aligned, 
               family = unwrap_von_mises,
               prior = prior_divergence, # the kappa and mu priors
               stanvars = stan_unwrap_fun + mod_circular_fun,
               silent = 2,
               backend = 'cmdstan'
)

#both models converge well with rhats [1.000, 1.002]
sm_truefalse = lapply(list(true = mod_true, false = mod_false), summary)
#their mu estimates are very close, differing by only 2°, with overlapping CI.
round( deg(
  sm_truefalse$false$fixed['Intercept',
                                    1:4]
  ) )

#add Leave-One-Out Cross-Validation
loo_true = loo(mod_true)
loo_false = loo(mod_false)
loo_compare(loo_true, loo_false)
#expected angle model has higher predictive power
```
Now the expected angle has somewhat higher predictive power, even though both find a similar mean direction (95% CI -7°–10° for `mod_false`).