---
title: "Circular Dataset Structures"
output: html_document
date: "`r Sys.Date()`"
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Details

> **AUTHOR**
>
> James Foster

> **DESCRIPTION**
>
> Examination of different dataset structures that may arise in the study of animal navigation.

> **INPUTS**
>
> `unwrap_functions.R`

> **OUTPUTS**
>
> Plotted results.

> **REFERENCES**
>
> *data types*
>
> Duelli P. & Wehner R.
> (1973) The Spectral Sensitivity of Polarized Light Orientation in Cataglyphis bicolor ( Formicidae , Hymenoptera ) *Journal of Comparative Physiology* 53(3) 37-53
>
> Edrich, W., Neumeyer, C.
> and von Helversen, O.
> (1979).
> “Anti-sun orientation” of bees with regard to a field of ultraviolet light.
> *J. Comp. Physiol.* 134, 151–157.
>
> *modelling methods*
>
> Sayin S, Couzin-Fuchs E, Petelski I, Günzel Y, Salahshour M, Lee CY, Graving JM, Li L, Deussen O, Sword GA, et al. (2025) The behavioral mechanisms governing collective motion in swarming locusts.
> Science.
> 387(6737):995–791
>
> Gabry J, Češnovar R, Johnson A (2022).
> cmdstanr: R Interface to 'CmdStan'.
> <https://mc-stan.org/cmdstanr/>
>
> Bürkner, P.-C.
> (2018).
> Advanced Bayesian Multilevel Modeling with the R Package brms.
> The R Journal 10, 395–411.
>
> Carpenter, B., Gelman, A., Hoffman, M. D., Lee, D., Goodrich, B., Betancourt, M., Brubaker, M., Guo, J., Li, P. and Riddell, A.
> (2017).
> Stan: A Probabilistic Programming Language.
> Journal of Statistical Software 76 doi: 10.18637/jss.v076.i01

> **TODO**
>
> -   move necessary functions to unwrap_functions.R
> -   make functions for result extraction from BRMS

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
